---
title: "Survivalanalysis using probabilistic models"
format: html
editor: visual
---

# Introduction

Let's download all the necessary libraries:

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(brms)
library(tidybayes)
library(ggplot2)
library(TreeSummarizedExperiment)
library(posterior)
library(dplyr)
library(tidyr)
library(stringr)
library(survival)
library(survminer)
library(vegan)
library(mia)
```


# Survival model
## Data manipulation

All data preprocessing was carried out in a separate script (`data.R`) to ensure clarity and modularity in the analysis workflow. The script loads the raw data, applies appropriate transformations, removes invalid samples, and derives variables needed for modeling.

To reduce dimensionality and avoid overfitting in probabilistic survival modeling, only the 5 most prevalent microbial taxa were selected based on their presence across samples. This targeted feature selection improves the robustness and interpretability of the model, especially given the relatively limited sample size.

```{r, include=FALSE}
# Download the data preprocessing
source("data.R")
```



## Modeling

We construct a Cox model formula using the log-transformed predictors already available in dff. A probabilistic Cox proportional hazards model is then fitted using brm(), with a normal(0, 1) prior on each bacterial coefficient to reflect weak prior knowledge. All of this is implemented in a separate script (`model.R`), and we now load the results.

```{r}
# Load model
fit_brms <- readRDS("fit_brms.rds")
```

We take summary of the model.

```{r}
# Print model summary
summary(fit_brms)
```


The model shows good convergence diagnostics. All Rhat values are equal to 1.00, indicating that the chains have converged well. Additionally, both Bulk and Tail Effective Sample Sizes (ESS) are high across all parameters, suggesting that the posterior distributions are well estimated and the sampling was efficient.

Posterior samples of the coefficients are extracted and summarized by computing the median, 2.5%, and 97.5% quantiles, forming the 95% credible interval for each predictor.

```{r, warning=FALSE}
# Extract raw draws as dataframe
draws_df <- as_draws_df(fit_brms)

# Select only b_ variables, excluding intercept
posterior_summary <- draws_df %>%
  select(starts_with("b_")) %>%
  select(-b_Intercept) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  group_by(variable) %>%
  summarise(
    median = median(value),
    lower = quantile(value, 0.025),
    upper = quantile(value, 0.975)
  ) %>%
  mutate(variable_clean = str_remove(variable, "b_"))


# Add significance column (if 95% CI does not include 0, (log(HR)=0))
posterior_summary <- posterior_summary %>%
  mutate(significant = ifelse(lower > 0 | upper < 0, "yes", "no"))
```


## Visualization

The exponentiated posterior medians and credible intervals are plotted as hazard ratios. Predictors whose intervals do not cross 1 (indicating statistical relevance) are highlighted in red, and others in black. The plot is sorted by effect size.

```{r}
# Plot hazard ratios
ggplot(posterior_summary, aes(y = reorder(variable_clean, median), 
                              x = exp(median), 
                              xmin = exp(lower), 
                              xmax = exp(upper), 
                              color = significant)) +
  geom_point() +
  geom_errorbarh(height = 0.2) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray") +
  scale_color_manual(values = c("yes" = "red", "no" = "black")) +
  labs(x = "Hazard Ratio", y = "", title = "Posterior estimates for all predictors (log-transformed)")

```

This figure displays the posterior estimates of hazard ratios for all log-transformed bacterial predictors. Each point represents the median of the posterior distribution, and the horizontal lines show the 95% credible intervals. Predictors highlighted in red are statistically significant, meaning their credible intervals do not include 1.

Hazard ratios below 1 (to the left of the dashed line) are associated with a reduced risk of mortality, while those above 1 suggest an increased risk. In this model, most predictors show no statistically significant association with mortality; only g_Bacteroides demonstrates a significant effect, indicating a potential reduce in risk.


## Survival curve

A fully probabilistic  approach to modeling the survival curve is not used here, as its implementation is technically complex and not readily supported by standard R packages. Instead, the classical Kaplan–Meier estimator is applied to provide a robust and interpretable nonparametric summary of cumulative mortality over time.

```{r}
# Fit Kaplan-Meier survival model
surv_fit <- survfit(Surv(Event_time, Event) ~ 1, data = dff)

# Plot cumulative mortality curve
ggsurvplot(
  surv_fit,
  data = dff,
  conf.int = TRUE,
  fun = "event",  
  palette = "blue",
  xlab = "Time (years)",
  ylab = "Cumulative mortality (%)",
  title = "Overall cumulative mortality"
)
```

This plot shows the overall cumulative mortality over a 10-year follow-up period. The curve represents the estimated probability of death over time in the entire cohort. The shaded area around the curve indicates the 95% confidence interval, and the small vertical ticks represent censored observations. By the end of the follow-up, cumulative mortality reaches approximately 60%, with a fairly steady increase over time.




